Our Members

### 2016-01-17 pnsac/newsletter/article_Feature1.tex
Adding and updating content for Jan/Feb 2016 newsletter.

\title{RAF Lightning with Over-Wing Tanks}
%\subtitle{Part 4}
\author{Jim Riddoch}

@@

During the early 60's Britain's Defence Department decided that the
new fleet of Lightning aircraft should have the capability to extend
its current sortie of 22 minutes to permit ferry to the Middle
East. In-flight refuelling had not worked successfully due to the
offset probe positioned under one wing. Accordingly BAC (British
Aircraft Corporation,Warton Division) was awarded a contract to
develop extended wing tanks that could be jettisoned if necessary.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{jim_riddoch.eps}
   %caption of the figure 
   \caption*{\small \em Jim Riddoch.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:jim_riddoch.eps}
\end{figure}

%% \begin{footnotesize}
%% [Image: {\normalfont\color{blue}\href{http://www.airteamimages.com/english-electric-lightning_XP693_united-kingdom---royal-air-force_43865.html}{English
%%     Electric Lightning with over-wing tanks}}]
%% \end{footnotesize}

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.5]{Eng3Intercooler.eps}
%   %caption of the figure 
%   \caption*{\small \em Garry installing the intercooler on engine 3.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:Eng3Intercooler}
%\end{figure}

Unfortunately the Lightning with its high speed wings had limited
capacity for wing tip tanks and already with armament pylons under
wing could not accommodate under-wing tanks, so the decision was made
to develop over-wing tanks secured with pylons through the wing
structure. These tanks would each carry 250 imperial gallons to
provide initial fuel transfer after take off and be jettisoned once
empty.

I was designated to be the Development Engineer to carry out the
development and testing of this design about 1963/64. The design
concept of these tanks had to include a balanced transfer system to
maintain aircraft C of G during take off and climb out, this meant 5
small interior sealed compartments pressurized with air to permit
sequential transfer. Also an emergency fuel jettison system prior to
tank wing ejection. As you can readily appreciate this was a very
challenging design and even more demanding for testing and
certification.

Once the initial test tanks were built they were dispatched to
Mechanical Test Department at Warton Lancashire, where I worked.
Needless to say the RAF were keen to witness the development and
testing of these over-wing tanks. A Wing Commander headed a team of
three Senior Techs to keep track of our progress. Fortunately the
Techs were very helpful and hands-on in working with us.  You can
imagine the challenges we faced checking the transfer sequences at
varying attitudes. Many a Saturday morning was spent testing and
verifying these change in attitudes and minimizing fuel transfer loss,
the end result being approximately 225 gallons of fuel transferred
from each wing tank depending on aircraft attitude and roll.

The next major design requirement was the emergency fuel jettison;
this was to be accomplished by air pressure from centre compartment to
rear of the tank where a jettisonable fairing was fastened by
explosive bolts and attached by a cord to a bung on the end of the
fuel ejection pipe. Talk about Heath Robinson device, this takes the
biscuit! Suffice to report that on the ground test of this
ill-conceived system with Chief Designer and my Chief Engineer and
many observers with cameras, the rear fairing departed into oblivion
and the cord broke, due to a miserable sub standard cord substituted
instead of a parachute cord as specified. The bung remained firmly
attached to the jettison pipe. It was about this time I first thought
of emigrating to Canada, in fact anywhere outside Britain!

Needless to say the cord was replaced, the ejection system proved
satisfactory and the tanks were certified for flight test. This was to
be done with one tank attached to the top of one wing and painted with
black and white checkered squares for easy recognition when ejected. I
always remember the Test Pilot, Peter Knight inquiring about the
likely occurrence of the tank separating from the wing and hitting the
tail of the aircraft. I assured him that ground testing of a model in
the wind tunnel had worked, in fact the small tank had disappeared
through the end of the tunnel and took some time to recover in the
neighbouring field.

To replicate the fuel system jettison some bright spark proposed
filling the tank with coloured water for better visibility. The
aircraft took off, circled Warton Aerodrome and made a low pass to
jettison the coloured water over the attentive crowd including chiefs
and dignitaries. Fortunately, the jettison of the tail fairing and
bung release worked perfectly and the water emerged in a colourful
shower.

The weather gods not being with us, the wind blew all the water over
the spectators covering their expensive clothing with a colourful dye!
Finally, the tank was ejected safely over Morecombe Bay and eventually
recovered and returned to BAC Warton amidst embarrassed and colourful
faces. As one might expect, this completely successful test left the
Wing Commander disgusted with the entire experiment. He cancelled the
project.

Any thoughts I had of joining the RAF completely disintegrated. Once
again, immigration to Canada looked much more inviting.

>>>>

\title{Volunteers Working on North Star Restoration}
%\subtitle{Part 4}
\author{Tom MacGregor, Legion Magazine}

@@

% \textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
%  Sqn, RCAF.}\\

\begin{quotation}
	You have to be careful opening up the panels covering the 
	engines, you never know if a bird or some animal might have made
	its home there.
\end{quotation}

That’s one lesson a group of dedicated volunteers has learned in its
10 years of painstaking work restoring a Royal Canadian Air Force
North Star aircraft at the Canada Aviation and Space Museum in Ottawa.

%\begin{figure*}[htbp]
%	\vspace{2em}
%	\centering
%	%name of the graphic, without the path AND in EPS format:
%	\includegraphics[scale=0.5]{op-hawk-one.jpg}
%	%caption of the figure 
%	%\caption*{\small \em Volunteers put in countless hours restoring a
%	North Star aircraft.}
%	%label of the figure, which has to correspond to \ref{}:
%	\label{fig:op-hawk-one.eps}
%\end{figure*}

The aircraft had sat neglected outside the museum for
years. “Essentially, the RCAF flew this one into the Rockcliffe Airport
when it was decommissioned. It sat in a old until Robert Holmgren said
someone should do something about it,” said project manager Bruce
Gemmill.

The airplane had been stripped of its non-essential gear and turned
over to what was then the National Aeronautical Collection, housed in
hangars built during the Second World War at the former RCAF station
at Rockcliffe in Ottawa. Time and weather took its toll. Birds and
animals made nests where they could get inside the panelling.

“We approached the museum about restoring the aircraft. At first, they
didn’t know how to take us,” said Gemmill. “The museum had its own
professional conservation staff but it had never worked with a team of
volunteers.”

Eventually the museum agreed to work with the volunteers who formed
the Project North Star Association of Canada with the late Robert
Holmgren as president. “We became the model for other volunteer groups
now working with the museum,” said the association’s current
president, Richard Lodge.

Now the aircraft gets towed into a hangar where it is available to the
volunteers between September and May before it is towed outside for
the summer.

The North Star was built by Canadair in the late 1940s and 1950s. It
was the RCAF version of the DC-4 civilian aircraft except that it used
four Rolls-Royce Merlin engines. Canadair built 71 of the aircraft.

The RCAF assigned the rst North Stars to 412 Squadron where they
transported VIPs and were used in various transfigurations for
reliable, long-range transport services.

During the Korean War the North Star was used by No. 426 (Thunderbird)
Sqdn. to ferry sup- plies across the Pacific Ocean to Japan (Operation
Hawk: The Korean Airlift, July/August). They would y 599 round trips
over the Pacific and deliver seven million pounds of cargo and 13,000
personnel on return trips. All this was achieved without a fatal
crash.

Most of the air force’s North Stars were declared surplus in the
1970s.

In the 10 years that Project North Star has been in operation, the
group has restored two of the four Merlin engines with a third one
nearly completed.

Each engine had to be taken o the aircraft, brought into the museum’s
shops where they were taken apart. Each item was tagged, catalogued 
and usually photographed. “We take a lot of photographs. That’s
how we know how they go back together,” said Lodge.

Engine project leader Garry Dupont said, “The first engine took us about
four and a half years to restore. We’re almost finished the third engine
which as taken us about two years. So we are getting faster.” The
propellers were restored separately by Hope Aero Propeller and
Components Inc. in Mississauga, Ont. All restoration work has to be up
to the professional standards set by the museum.

The group plans to restore the aircraft to its original look, not to
have it y again.

Lodge said that they have some RCAF manuals but a lot of time they do
not help when it comes to putting 50-year-old parts together today.

The aircraft itself, 17515, has seen countless hours of work put into
re- storing the cockpit with all its switches and navigation
equipment and a small galley for the crew of seven.

Still to be done is work in the cargo section and painting the
exterior of the aircraft.

“We know this aircraft ew during the Korean War. In fact, we have 
pictures showing it configured with litters to bring the wounded back to
Canada,” said Gimmell. “We don’t want to put in the litters but we
have people working on creating the rope seats that were used.”

The plan is to paint the aircraft to its post-1965 markings, since it
was painted with the Maple Leaf ag on it.

Still the search goes on to nd new members for the association. “The
important thing is to nd new people to whom we can pass on this 
knowledge and skills we are developing,” said Lodge.

### 2016-01-17 pnsac/newsletter/article_MemberInterview.tex
Adding and updating content for Jan/Feb 2016 newsletter.

\title{North Star Merchandise}
\author{Phil Chrysler}

@@

The idea for North Star merchandise started with a need to identify those volunteers that were assisting when the aircraft was put on display to the general public.  That requirement evolved into the blue member's only crested golf shirts (men and ladies).  Over time we have added baseball caps, toques sweatshirts, T-shirts and fleece vests (by special order) all with the North Star crest.  Also a 1:200 scale model of the aircraft was produced in limited numbers of which we still have a few.  With the exception of the members only golf shirt all the other items are available at the Aviation and Space Museum store.  The items are available to members of the North Star directly at a discount.

We also have a CD of the North Star operating manual and a North Star pin that are not available in the Museum store.

From the beginning our purpose was to bring more awareness to the general public of our restoration activities rather generate money on the sale of the items.  So we have made an effort to keep the quality high and profits modest.  All money earned is used to advance the restoration effort.  The North Star web site has prices and pictures of the available items.

Merchandise page: {\normalfont\color{blue}\texttt{\url{http://www.projectnorthstar.ca/menu-merchanidise}}}

>>>>

\title{Our Members}
\author{Interview with Bruce Gemmill}

@@

\textit{Bruce Gemmill has been involved with Project North Star since 2005 and
	has exceeded 9,000 volunteer hours. Bruce is presently the Project
	Manager, as well as a member of the board of directors.}
